---
output:
  html_document:
    code_folding: hide
    css: assets/style.css
    self-contained: no
---
# CHICAGO RATS BE GONE
<p class="text-center subtitle-size">Visualizing Chicago's rats infestation and sanitation problems</p>
<hr class="divider">
```{r error=FALSE, warning=FALSE, results='hide', message=FALSE}
# Libraries
library(lubridate)
library(tidyverse)
library(dplyr)
library(gghighlight)
library(ggrepel)
library(ggthemes)
library(readxl)
library(scales)
library(statar)
library(stringr)
library(tidyr)
library(here)
library(httr)
library(jsonlite)
library(magrittr)
```

## Emerging Rat Concerns...
In recent years, Chicago has been receiving increasing numbers of rodent infestation complaints. It is important for the city government to actively engage in resolving rat complaints, as rats can cause huge property damage and spread harmful pathogens to humans. This exploratory visualization project illustrates the severity and patterns of rat complaints within the city of Chicago, and seeks to help policymakers to find effective measures to mitigate rodent problems.
```{r cache=TRUE, error=FALSE, warning=FALSE, results='hide', message=FALSE}
# Reading in Data
census_tracts <- 
  read_csv(here("data", 
                "tract_community.csv"))

community_numbers <-
  read_excel(here("data", 
                  "community_numbers.xlsx"))

chicago_rodents <- 
  read_csv(here("data", 
                "Chicago_311_Rodent_2014-2018.csv"))

boston_311 <- 
  read_csv(here("data",
                "Boston_311.csv"))

dc_311_2018 <- 
  read_csv(here("data",
                "DC_Service_Requests_2018.csv"))

nyc_rodents_2018 <- 
  read_csv(here("data",
                "NYC_311_Rodent_2018.csv"))

la_rodents_2018 <- 
  read_csv(here("data",
                "LA_Dead_Animal_Removal_2018.csv"))

detroit_rodents_2018 <- 
  read_csv(here("data",
                "Detroit_Rodent_2018.csv"))

# Scraping Data from ACS
base <- "https://api.census.gov/data/2016/acs/acs5?"
acs_vars <- "get=B00001_001E,B07013_001E,B07013_003E,B19301_001E"
state_tracts <- "&for=tract:*&in=state:17"
AUTH_TOKEN <- 
  read_delim(here("data", "Auth.txt"), 
             delim = "/", 
         col_names = F)$X1
key <- paste("&key=", AUTH_TOKEN, sep = "")

url <- paste(base, acs_vars, state_tracts, key, sep = "")
response <- GET(url)
all_il_tracts <- 
  jsonlite::fromJSON(content(response, as = "text")) %>%
  data.frame() %>%
  slice(-1)

colnames(all_il_tracts) <- 
  c("Population",
    "Total_Living_In_Area",
    "Renter_Count",
    "Income_Per_Capita",
    "State",
    "County",
    "Tract")
```

## Understand Rat Problems through Graphs
```{r error=FALSE, warning=FALSE, results='hide', message=FALSE}
# Preparing Data
census_tracts <-
  census_tracts %>%
  left_join(community_numbers, 
            by = c("Community Area" = "Number"))

chicago_rodents_2018 <-
  chicago_rodents %>%
  filter(year(mdy(`Creation Date`)) == 2018)

boston_rodents_2018 <-
  boston_311 %>%
  filter(str_detect(TYPE, "Rodent"),
         year(open_dt) == 2018)

dc_rodents_2018 <-
  dc_311_2018 %>%
  filter(str_detect(SERVICECODEDESCRIPTION, "Rodent"))

chicago_rat_community_17 <-
  chicago_rodents %>%
  filter(year(mdy(`Creation Date`)) == 2017) %>%
  group_by(`Community Area`) %>%
  summarise(`Number of Complaints` = n()) %>%
  inner_join(community_numbers, by = c("Community Area" = "Number")) %>%
  select(Community, `Number of Complaints`)

to_numeric <- 
  function(column) {
    return(as.numeric(levels(column))[column])
    }

Chicago_communities <- 
  all_il_tracts %>%
  mutate(Census_Tract = str_c(State, County, Tract),
         Total_Income = to_numeric(Income_Per_Capita) * 
                        to_numeric(Population)) %>%
  select(Census_Tract, 
         Population,
         Total_Income,
         Total_Living_In_Area, 
         Renter_Count) %>%
  mutate(Census_Tract = as.numeric(Census_Tract)) %>%
  inner_join(census_tracts, 
             by = c("Census_Tract" = "Census Tract")) %>%
  mutate_at(vars(-"Community",
                 -"Total_Income"), 
            .funs = "to_numeric") %>%
  select(-Census_Tract, -`Community Area`)

Chicago_communities <- 
  Chicago_communities %>%
  group_by(Community) %>%
  summarise_all(.funs = sum) %>%
  mutate(`Income Per Capita` = round(Total_Income / Population, 0),
         `Renter Proportion` = Renter_Count / Total_Living_In_Area) %>%
  select(Community, Population, `Renter Proportion`, `Income Per Capita`)

chicago_rat_community_17 <-
  Chicago_communities %>%
  inner_join(chicago_rat_community_17, by = "Community")

chicago_rat_community_17 %<>%
  mutate(`Number of Requests per 10000 People` = 
           round(`Number of Complaints` / Population * 10000, 0)
         )
```

### "Rat Capital" of United States?
In 2018, we saw a total of 37577 rat complaints in Chicago. Compared with other major US cities like New York and Los Angeles and cities that were historically troubled by rodent infestations such as Detroit, Chicago appeared to have the largest number of complaints all year round. From the graph below, we can also see that rat problems were especially prevalent in the summer.
```{r error=FALSE, fig.height=7., fig.width=11, message=FALSE, warning=FALSE}
# Data Processing for Plot 1
city_mon_summary <- 
  function(df, date_col, city_name) {
    monthly_summary <-
      df %>%
      group_by(month(date_col, 
                   label = TRUE, 
                    abbr = TRUE)) %>%
      summarise(`Number of Complaints` = n(),
                                  City = city_name) %>%
      rename(Month = 1)
    return(monthly_summary)
    }

chicago_months <- 
  city_mon_summary(chicago_rodents_2018,
                   mdy(chicago_rodents_2018$`Creation Date`),
                   "Chicago")

boston_months <- 
  city_mon_summary(boston_rodents_2018,
                   boston_rodents_2018$open_dt,
                   "Boston")

dc_months <- 
  city_mon_summary(dc_rodents_2018,
                   dc_rodents_2018$ADDDATE,
                   "Washington DC")

nyc_months <- 
  city_mon_summary(nyc_rodents_2018,
                   mdy_hms(nyc_rodents_2018$`Created Date`),
                   "New York City")

la_months <-
  city_mon_summary(la_rodents_2018,
                   mdy_hms(la_rodents_2018$CreatedDate),
                   "Los Angeles")

detroit_months <- 
  city_mon_summary(detroit_rodents_2018,
                   mdy_hms(detroit_rodents_2018$`Created At`),
                   "Detroit")

city_months_18 <- 
  bind_rows(chicago_months,
            boston_months,
            dc_months,
            nyc_months,
            la_months,
            detroit_months)

# Plot 1
city_months_18 %>%
  ggplot(aes(x = Month, 
             y = `Number of Complaints`, 
         group = City,
         color = City)) +
  geom_point() +
  geom_line() + 
  geom_line(data = chicago_months,
            size = 1.1) +
  labs(
           title = "Rat Complaints by Highly Affected US Cities in 2018",
        subtitle = "Chicago generally has the most rat complaints all year round",
         caption = "\nSource: City Open Data Portals"
    ) +
  scale_color_manual(values = c("#098154", 
                                "#c72e29",
                                "#016392", 
                                "#be9c2e", 
                                "#fb832d",
                                "#000000")) +
  theme_wsj(
       base_family = "Times New Roman", 
      title_family = "Times New Roman"
      ) +
  theme(
        plot.title = element_text(size = 16,
                                 hjust = 0.42),
     plot.subtitle = element_text(size = 14,
                                 hjust = 0.42),
      plot.caption = element_text(size = 10),
                plot.margin = margin(t = 30, 
                                     r = 30, 
                                     b = 30, 
                                     l = 30, 
                                  unit = "pt"),
        axis.title = element_text(size = 13,
                                  face = "bold"),
       legend.text = element_text(size = 13)
     ) +
  guides(color = guide_legend(title = NULL))
```

### Community Analysis
Although rat complaint distribution is not equivalent of the actual rat distribution, Chicago Lincoln Park Zoo has recently conducted <a href="https://www.frontiersin.org/articles/10.3389/fevo.2018.00189/full" target="_blank">research</a> indicating that it can be highly reflective of rodent infestation situation. That begs the question what kind of communities are more likely to complain about rats. Looking at the data of 2017, we can see that lower-income communities with less renters are less likely to complain about rats. It appears counter-intuitive, but in reality lower-income renters tend to worry less about rats than other socioeconomic aspects of their lives.
```{r error=FALSE, fig.height=7., fig.width=11, message=FALSE, warning=FALSE}
# Data Processing for Plot 2
more_renter_low_income <-
  chicago_rat_community_17 %>%
  filter(`Income Per Capita` <= 21708) %>%
  mutate(`Income Level` = "Communities\nunder 180% FPL\n(Federal Poverty Level)")
  
# Plot 2
chicago_rat_community_17 %>%
  ggplot(aes(x = `Renter Proportion`,
             y = `Number of Requests per 10000 People`)) +
  geom_point(aes(size = `Income Per Capita`), 
                alpha = 0.8, 
                color = "gray") +
  geom_point(data = more_renter_low_income,
         aes(size = `Income Per Capita`, 
            color = `Income Level`)) +
  geom_smooth(data = more_renter_low_income, 
             color = "dodgerblue4",
                se = F,
      aes(linetype = "For Lower Income \nCommunties, the Larger\nRenter Proportion, \nthe Less Complaints"),
              size = 1.1) +
  labs(
    title = "Lower-income Communities with More Renters Are Less \nLikely to Complain about Rats in Chicago (2017)",
 subtitle = "Lower-income renters may care more about \nother socioeconomic issues than rat problems",
  caption = "Source: Chicago Data Portal & American Community Survey",
        x = "Proportion of People Renting in Community",
        y = "Number of Rodent Complaints"
    ) +
  scale_color_wsj("colors6", "Income Level") +
  scale_size(breaks = c(20000, 40000, 60000, 80000),
             labels = c("20000 US dollars",
                        "40000 US dollars",
                        "60000 US dollars",
                        "80000 US dollars")) +
  scale_linetype_manual(values = c("dashed", "dotted")) +
    theme_wsj(
       base_family = "Times New Roman", 
      title_family = "Times New Roman"
      ) +
  theme(plot.title = element_text(size = 16,
                                 hjust = 0.4),
     plot.subtitle = element_text(size = 14,
                                 hjust = 0.4),
      plot.caption = element_text(size = 10),
       plot.margin = margin(t = 30, 
                            r = 30, 
                            b = 30, 
                            l = 30, 
                         unit = "pt"),
       legend.text = element_text(size = 10),
      legend.title = element_text(size = 12, 
                                  face = "bold"),
  legend.direction = "vertical",
   legend.position = "right",
        axis.title = element_text(size = 13, 
                                  face = "bold")) +
   guides(linetype = guide_legend("Community Trend"))
```

